---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wombat21 <img src= "https://monasharp.github.io/game-package-wombats/logo.png" align="right" height="139" alt="" />

<!-- badges: start -->

[![R-CMD-check](https://github.com/MonashARP/game-package-wombats/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/MonashARP/game-package-wombats/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## What is wombat21?

**wombat21** is an R package that simulates a fully playable game of Blackjack (21). It supports:

-    Multiplayer mode with up to 6 players (human or AI)

-    Customizable dealer logic with automation support

-    Betting system with coin tracking

-    Game logic: splitting, doubling down, and insurance

-    Strategy simulation with user-defined rules

-    Bankroll visualization across rounds

-    Auto-save and resume for player profiles

## Installation

You can install the development version of wombat21 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("MonashARP/game-package-wombats")
```

Then load the package:

``` r
library(wombat21)
```

## Quick Start

```{r eval=FALSE}
library(wombat21)

# Start a multiplayer Blackjack game (interactive mode)
play()

# Or simulate 1000 single-player strategy games (simulation mode)
simulate_blackjack(num_rounds = 1000)
```

To explore saved player data, use:

```{r eval=FALSE}
get_home_dir()
list.files(get_home_dir())
```

## Gameplay Mode: Interactive Blackjack

The play() function launches an interactive game for 1–6 players. Features include:

- Mix of human and AI players(users can choose how many AI or human players wanted)

- Full support for Blackjack rules: hit, stand, double down, split, insurance

- Dealer plays automatically using hard-coded logic(user can customize dealer logic if desired)

- Coin balances persist across rounds(initial amout 1000 coins, minimum 10 coins bnets per game, auto bankrupt detection)

- Game runs entirely in the R console

Gameplay loop:

1. Enter player names and bets

2. Deal cards

3. Each player takes turns

4. Dealer plays automatically

5. Results are calculated and coins updated

Example:

```{r eval=FALSE}
# Start a 2-player game: 1 human, 1 AI
play()

# Example gameplay prompt:
# > Enter total number of players (1–6): 2
# > Enter number of computer players (0–2): 1
# > Enter unique name for Human Player 1: alice 
# > 
# > New player created: alice with 1000 coins.
# > New computer player created: Computer_1 with 1000 coins.
# > 
# > alice, enter your bet (available: 1000): 10 
# >                       // Here,You can place bets; assume 10 coins now
# > Computer_1 (computer) bets 10 coins.
# > 
# > alice's hand: K♠ Q♣   |  Bet: 10
# > Computer_1's hand: 3♣ ?   |  Bet: 10
# > Dealer shows: 8♦ ?
# > 
# > --- alice ('human') Playing ---
# > Hand 1: [K♠ Q♣]
# > Score: 20
# > Hit, Stand or Double? (hit/stand/double/exit):

# After making your move, the game proceeds to the dealer's turn, outcomes are revealed, and coin balances are updated automatically.
```

### Customizing Dealer Logic

By default, the dealer hits until reaching 17 or more, like standard Blackjack.
To override this behavior, you can supply a custom function in the simulation mode (not `play()` yet):

```{r eval=FALSE}
# Custom dealer that stops at 16 instead of 17
custom_dealer <- function(hand) {
  score <- calculate_score(hand)
  if (score < 16) return("hit")
  else return("stand")
}
```

In the current release, `play()` uses hard-coded dealer logic, but simulation (`simulate_multiplayer_blackjack()`) allows for pluggable decision functions.

## Simulation Mode: Strategy Testing

Use `simulate_multiplayer_blackjack()` to test decision strategies against a standard dealer.

Supports:

- Custom player strategies for split or insurance (pass functions)

- Repeated rounds with reproducible outcomes

- Summary of win/loss stats

```{r eval=FALSE}
simulate_blackjack(
  num_rounds = 500,
  split = TRUE,
  buy_insurance = function(hand, dealer_card) dealer_card == "A"
)
# result checking
sim$win_rate
sim$split_rate
sim$final_coins
```

Simulation is **single-player** only, designed for strategy evaluation—not interactive gameplay. Make sure to set **enough initial coins** for large simulations.

## Game Data Management

### Saving Game Data

Player data (names, coin balances, history) are stored locally using `get_home_dir()`. Unique player names are enforced.

- Coin balances and player profiles are saved between sessions

- Unique names prevent overwriting

- Use `unlink(get_home_dir(), recursive = TRUE)` to reset all

```{r eval=FALSE}
get_home_dir()
#> "/Users/yourname/.wombat21"

list.files(get_home_dir())
#> [1] "players.rds" "bankroll.rds"
```

### Reset Game Data

To remove all saved progress and start fresh:

```{r eval=FALSE}
# Permanently delete all saved players and history
unlink(get_home_dir(), recursive = TRUE)
```

This will reset the game to its default state. New players will be prompted to enter their names and start with 1000 coins.

## Results Visualization

You can use `plot_bankroll_history()` to visualize how each player's coin balance changes over time:

```{r eval=FALSE}
# Example: plot bankrolls from a previous game session
bankroll_data <- load_players()$bankroll_history

plot_bankroll_history(bankroll_data)
```

The plot shows the total coins for each player across rounds, helping users evaluate performance and strategy over time.

## Learn More

To learn more about how each function works, you can view the documentation directly in your R console:

```{r eval=FALSE}
?<function_name>
# e.g.
?simulate_multiplayer_blackjack
?plot_bankroll_history
```

Advanced users can also explore internal helper functions by browsing the source code or calling them directly with the ::: operator:

```{r eval=FALSE}
# Call an internal function (use with care)
wombat21:::play_dealer_turn
```

For a full overview of the package, check out the vignettes and function references:

👉 Package Website (Documentation): 

- [Get Started guide](https://monasharp.github.io/game-package-wombats/articles/wombat21.html) – a walkthrough of basic usage and gameplay.

- [Reference tab](https://monasharp.github.io/game-package-wombats/reference/index.html) – full API documentation for all user-facing functions.

## 💡Have a try now ~

We hope you enjoy exploring strategies, simulations, and game logic with wombat21!

By now, you probably have a pretty good idea of what this package is all about — and it’s time to put that knowledge into play.

Happy coding, and enjoy the game.🍀
