---
title: "Advanced Features and Design of wombat21"
author: "Sun Yan Joanna Ma, Wei Zhu Chen, Aneesh Agarwal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Features and Design of wombat21}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 🧩 Object System with vctrs

### **Why Use vctrs?**

The `vctrs` package provides a robust foundation for type-safe, extensible object-oriented programming in R. In **`wombat21`**, it powers the entire game logic for cards and hands.

**1. Custom Card and Hand Classes**

`wombat21` defines:

-   **`card`**: an object with `rank` and `suit`, e.g., `"A♠"`, `"10♦"`

-   **`blackjack_hand`**: a player hand consisting of `card` objects

Both are built with `vctrs::new_rcrd()`, enabling compact storage and automatic validation.

```{r eval=FALSE}
card <- function(rank, suit) {
  vctrs::new_rcrd(list(rank = rank, suit = suit), class = "card")
}

new_blackjack_hand <- function(cards) {
  vctrs::new_rcrd(list(cards = cards), class = "blackjack_hand")
}
```

**2. Helper Functions and Safe Manipulation**

Custom card and hand objects can be safely manipulated using helper functions:

```{r eval=FALSE}
# Create cards and a hand
c1 <- card("A", "♠")
c2 <- card("10", "♦")
hand <- new_blackjack_hand(c(c1, c2))

card_rank(hand$cards)    # Returns c("A", "10")
card_value(hand$cards)   # Returns c(11, 10)
```

*These functions ensure your game logic is **clean**, **type-safe**, and **readable**.*

**3. Extending with S3 + vctrs**

All relevant behaviors—printing, type coercion, equality—are registered explicitly via `vctrs::s3_register()`.

```{r eval=FALSE}
# Custom print method for card objects
print.card <- function(x, ...) {
  cat(sprintf("<card: %s of %s>\n", card_rank(x), card_suit(x)))
}

# Register vctrs casting and type methods for cards
vctrs::s3_register("vctrs::vec_cast", "card.character")
vctrs::s3_register("vctrs::vec_cast", "character.card")
vctrs::s3_register("vctrs::vec_ptype2", "card.character")
vctrs::s3_register("vctrs::vec_ptype2", "character.card")
vctrs::s3_register("vctrs::vec_ptype2", "card.card")
```

*This design ensures smooth interoperation with base R vectors and custom behaviors.*

### Benefits for Advanced Users & Developers

-   **Safer code:** Bugs from invalid card or hand objects are caught early

-   **Clearer APIs:** Functions clearly expect and return game objects

-   **Easy extension:** Add new rules or card types with minimal refactoring

-   **Better debugging:** Informative errors when misuse is detected

## 🧵 Metaprogramming & Generic Interfaces

While **`wombat21`** does not heavily rely on *metaprogramming*, it demonstrates simple use of **expression evaluation** and **S3 generics** to enable flexible logic and modular game components.

### Metaprogramming in Action

A notable use of metaprogramming appears in our AI hit/stand decision logic:

```{r eval=FALSE}
while (rlang::eval_tidy(rlang::expr(!!current_score < 17))) {
  ...
}
```

This pattern lets us **evaluate conditional logic dynamically**, which could be extended to support more advanced decision trees or strategy DSLs.

Although simple here, this approach allows:

-   Dynamic game strategies (e.g., thresholds that are formula-based)

-   Pluggable control flows without hardcoding

-   Compatibility with user-defined logic in future extensions

### Interface Design via S3

Beyond that, **interface extensibility** in `wombat21` is achieved through clean **S3 generics**:

-   `card_rank()`, `card_value()`, and `calculate_score()` all support **class dispatch**

-   Functions like `card_rank.card()` and `card_rank.character()` make logic reusable across both S3 `card` objects and plain characters

This gives:

-   **Clear separation** between internal logic and user-visible objects

-   **Easy extension** of new types (e.g., `joker_card`, `double_card`)

-   Consistent API behavior with helpful fallbacks and method dispatch

### Summary

+---------------------+--------------------------+---------------------------------------------+
| Aspect              | Technique Used           | Purpose                                     |
+=====================+==========================+=============================================+
| Hit/stand condition | `rlang::eval_tidy()`     | Dynamically evaluate game thresholds        |
+---------------------+--------------------------+---------------------------------------------+
| Rank/value logic    | S3 generic dispatch      | Allow multiple input types (`card`, `char`) |
+---------------------+--------------------------+---------------------------------------------+
| Extendability       | `vctrs::new_rcrd()` + S3 | Build and validate extensible game types    |
+---------------------+--------------------------+---------------------------------------------+

### Want to do more?

-   For more advanced *metaprogramming*, consider:

-   User-defined strategy expressions via quosures

-   Parameterized hit/stand rules embedded in `play()` flow

-   Callable logic passed in as closures (e.g., `split_func`, `insurance_func` – which you already support)

## ⚡ Boosting Performance with Rcpp

To speed up critical logic like **score calculation**, `wombat21` uses a C++ backend via `Rcpp`. This ensures fast, reliable, and vectorized performance even during **high-volume simulation**.

### What’s Accelerated

The most computationally intense operation in Blackjack is score calculation—especially due to the **variable value of Aces**. Instead of handling this in pure R every round, we offload it to C++:

```{r eval=FALSE}
calculate_score(hand)
```

internally calls:

```{r eval=FALSE}
calculate_score_cpp(ranks)
```

### How It Works

```{r eval=FALSE}
# R wrapper (R/score_cal.R)
calculate_score <- function(hand) {
  ...
  ranks <- card_rank(cards)
  calculate_score_cpp(ranks)
}
```

```{r cpp-example, eval=FALSE}
# C++: src/calculate_score.cpp
# int calculate_score_cpp(std::vector<std::string> ranks);
```

This ensures:

-   **Low latency** even with thousands of hands

-   **Correct Ace handling** (e.g., dual 1/11)

-   **Seamless integration** with S3/vctrs objects

### When to Use

You don’t need to call the C++ version manually—`calculate_score()` handles this for you. But advanced users can test or use it directly:

```{r eval=FALSE}
calculate_score_cpp(c("A", "A", "9"))  # Returns 21
```

**Speed Comparison (10,000 hands)**

| Method                       | Elapsed Time (sec) | Relative Speed |
|------------------------------|--------------------|----------------|
| Pure R (`calculate_score_r`) | 0.28               | 1× (baseline)  |
| Rcpp C++ (`calculate_score`) | 0.06               | 4.7× faster    |

## 🧑‍💻 Advanced: Custom Strategy Functions

You can pass any function to the `split` or `buy_insurance` argument in `simulate_blackjack()`, enabling sophisticated, context-dependent strategies.

### 1. Advanced logic: Take dealer's upcard into account for split

```{r eval=FALSE}
# Only split A or 8, and never split if dealer shows a 10 or Ace.
split_fn <- function(hand, dealer_hand) {
  ranks <- card_rank(hand$cards)
  dealer_rank <- card_rank(dealer_hand$cards[1])
  length(ranks) == 2 &&
    ranks[1] == ranks[2] &&
    ranks[1] %in% c("A", "8") &&
    !dealer_rank %in% c("10", "A")
}
```

### 2. Advanced insurance logic

```{r eval=FALSE}
# Only buy insurance if player has blackjack AND dealer shows Ace.
insurance_fn <- function(player, dealer_hand) {
  is_blackjack(player$hand) && card_rank(dealer_hand$cards[1]) == "A"
}
```

### 3. Compare Strategies with Batch Simulation and Visualization

```{r eval=FALSE}
res_advanced <- simulate_blackjack(
  num_rounds = 10000,
  split = split_fn,
  buy_insurance = insurance_fn
)

res_simple <- simulate_blackjack(
  num_rounds = 10000,
  split = FALSE,
  buy_insurance = FALSE
)

barplot(
  c(Advanced = res_advanced$win_rate, Simple = res_simple$win_rate),
  beside = TRUE, col = c("skyblue", "grey"),
  main = "Win Rate Comparison: Advanced vs Simple Strategy"
)
```

*These examples show how you can build, test, and visualize sophisticated blackjack strategies by passing custom logic into the simulation functions.*

💡Try experimenting with your own logic—`wombat21` is designed to make strategy research both powerful and fun!

## 🧑‍🤝‍🧑 Team Contribution Summary

+----------------+-----------------------------------------------------------------------------------------+
| Name           | Key Contributions                                                                       |
+================+=========================================================================================+
| Weizhu Chen    | \- Designed and implemented main game loop and core mechanics\                          |
|                | - Developed S3 generics/methods (e.g., `card_rank`, `vec_cast`)\                        |
|                | - Modularized and refactored codebase\                                                  |
|                | - Added simulation feature and usage examples\                                          |
|                | - Improved documentation structure and naming consistency                               |
+----------------+-----------------------------------------------------------------------------------------+
| Joanna Ma      | \- Developed multi-player and computer player support\                                  |
|                | - Implemented betting and coin management\                                              |
|                | - Created visualization functions for player ranking and coin history\                  |
|                | - Debugged and enhanced S3/vctrs compatibility\                                         |
|                | - Helped with data storage and user interface troubleshooting                           |
+----------------+-----------------------------------------------------------------------------------------+
| Aneesh Agarwal | \- Developed and integrated C++ backend for fast hand scoring (`calculate_score_cpp`)\  |
|                | - Provided function examples and unit tests\                                            |
|                | - Contributed to vignette and design documentation (including metaprogramming choices)\ |
|                | - Assisted in troubleshooting and code review                                           |
+----------------+-----------------------------------------------------------------------------------------+

> For full details of team member roles, see the [project repository](https://github.com/MonashARP/game-package-wombats).
