---
title: "Advanced Features and Design of wombat21"
author: "Sun Yan Joanna Ma, Wei Zhu Chen, Aneesh Agarwal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Features and Design of wombat21}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 🎯 Design Philosophy

**wombat21** is built for three main usage modes—**interactive gameplay**, **batch simulation**, and **advanced visualization**—all with a modular, extensible, and high-performance architecture.

-   **Interactive Play:**\
    All user-facing gameplay logic is modular and transparent.\
    Players can interact with human or AI opponents, with clear APIs for betting, splitting, insurance, and persistent progress tracking.

-   **Flexible Simulation:**\
    The simulation engine is fully programmable:\
    Users can run thousands of games automatically, passing custom strategy functions (for splitting, insurance, etc.) and exploring the effects of advanced rule variations or AI behavior.

-   **Rich Visualization:**\
    Game results, player history, and simulation outcomes can be visualized interactively—making it easy to analyze trends, compare strategies, and communicate findings.

**Modularity, extensibility, and performance** are at the core of wombat21’s design.\
All core components (cards, hands, strategies, players) are built using robust object systems (vctrs, S3), and the scoring logic is accelerated with C++ for fast, large-scale analysis.

This design philosophy ensures wombat21 is not just a game, but a flexible engine for research, teaching, and advanced experimentation in R.

## 🏷️ Function Names and Arguments

All user-facing functions in `wombat21` are named to be clear, concise, and consistent, following common R conventions:

-   **Verb-based names:**\
    Functions perform an action and are named accordingly (e.g., `play()`, `simulate_blackjack()`, `calculate_score()`).

-   **Explicit parameter names:**\
    Arguments are descriptive, self-explanatory, and use snake_case (e.g., `num_rounds`, `bet_amount`, `buy_insurance`, `split`).

-   **Strategy customization:**\
    Where applicable, functions accept logical values or user-defined functions as parameters (`split`, `buy_insurance`), supporting flexible and advanced gameplay.

-   **Consistent interface:**\
    The same function can be used interactively (`play()`), for simulation (`simulate_blackjack()`), or for visualization (`plot_bankroll_history()`), all using familiar argument conventions.

-   **Object arguments:**\
    Game logic functions expect well-defined objects (e.g., `blackjack_hand`, `card`), ensuring type safety and clear data flow.

**Internal functions** (not exported) are named with clear, specific purposes, and marked with `@noRd` and `@keywords internal` (e.g., `generate_blackjack_hands()`, `calculate_score_r()`). These utilities support testing, benchmarking, or code reuse, but are not part of the main user API.

This consistent naming and argument style makes the package intuitive to use and easy to extend for both new and advanced users.

## 🗂️ Functions Specification

1.  **Card Objects:** Card validation and attribute extraction [*Main Arguments*: `card` ]

    -   [*card_utils.R*]{style="color:gray"} ：`card_is_face()`, `card_is_ace()`, `card_value()`, `card_suit()`, `card_rank()`

    -   [*card-class.R*]{style="color:gray"} ：`card()`, `format.card()`, `print.card()`, `vec_cast.card()`, `as.character.card()`

2.  **Hand Objects:** Hand creation and validation [*Main Arguments*: `cards`, `hand`]

    -   [*hand_utils.R*]{style="color:gray"} ：`new_blackjack_hand()`

    -   [*hand_type.R*]{style="color:gray"} ：`is_five_card_charlie()`, `is_blackjack()`

3.  **Scoring Logic:** Hand scoring and rules [*Main Arguments*: `hand`, `card`]

    -   [*score_cal.R*]{style="color:gray"} ：`calculate_score()`

    -   [*RcppExports.R*]{style="color:gray"} ：`calculate_score_cpp()`

4.  **Dealing & Setup:** Initial hands and deck operations [*Main Arguments*: `players`, `deck`]

    -   [*deal_card.R*]{style="color:gray"} ：`deal_cards()`

    -   [*setup.R*]{style="color:gray"} ：`setup_and_display_initial()`

5.  **Gameplay:** Main loop and player actions [*Main Arguments*: none]

    -   [*game_loop.R*]{style="color:gray"} ：`play()`, `ask_play_again()`, `exit()`

    -   [*player_logic.R*]{style="color:gray"} ：`player_action()`, `play_player_turns()`

6.  **Betting and Insurance** [*Main Arguments*: `players`, `hand`]

    -   [*bets.R*]{style="color:gray"} ：`get_bets()`

    -   [*Insurance.R*]{style="color:gray"} ：`handle_insurance()`

7.  **Splitting** [*Main Arguments*: `hand`, `players`]

    -   [*split.R*]{style="color:gray"} ：`handle_splitting()`

8.  **Simulation** [*Main Arguments*: `num_rounds`, `split`, `buy_insurance`]

    -   [*simulation.R*]{style="color:gray"} ：`simulate_blackjack()`

9.  **Result Judgement & Summary** [*Main Arguments*: `scores`, `players`, `hands`]

    -   [*results.R*]{style="color:gray"} ：`end_round()`, `display_final_results()`, `determine_winner()`

10. **Player I/O & Persistences** [*Main Arguments*: `players`, `bankroll`, `id_map`]

    -   [*input_player.R*]{style="color:gray"} ：`input_players()`

    -   [*io.R*]{style="color:gray"} ：`save_players()`, `load_players()`, `reset_game_data()`

11. **Visualization** [*Main Arguments*: `bankroll_history`]

    -   [*bankroll_graph.R*]{style="color:gray"} ：`plot_bankroll_history()`

    -   [*results.R*]{style="color:gray"} ：`plot_player_ranking()`

12. **Utilities** [*Main Arguments*: none]

    -   [*data_directory.R*]{style="color:gray"} ：\`get_home_dir()

## 🗃️ Core Data Structures

-   **card**: Represents a single playing card (fields: `rank`, `suit`)
-   **blackjack_hand**: Represents a player's hand (field: `cards`, a vector of card objects)
-   **players**: List of player objects (fields: `name`, `coins`, `is_computer`, ...)

This structure supports flexible extension, validation, and efficient manipulation via vctrs and S3.

## 🧩 Object System with vctrs

### **Why Use vctrs?**

The `vctrs` package provides a robust foundation for type-safe, extensible object-oriented programming in R. In **`wombat21`**, it powers the entire game logic for cards and hands.

**1. Custom Card and Hand Classes**

`wombat21` defines:

-   **`card`**: an object with `rank` and `suit`, e.g., `"A♠"`, `"10♦"`

-   **`blackjack_hand`**: a player hand consisting of `card` objects

Both are built with `vctrs::new_rcrd()`, enabling compact storage and automatic validation.

```{r eval=FALSE}
card <- function(rank, suit) {
  vctrs::new_rcrd(list(rank = rank, suit = suit), class = "card")
}

new_blackjack_hand <- function(cards) {
  vctrs::new_rcrd(list(cards = cards), class = "blackjack_hand")
}
```

**2. Helper Functions and Safe Manipulation**

Custom card and hand objects can be safely manipulated using helper functions:

```{r eval=FALSE}
# Create cards and a hand
c1 <- card("A", "♠")
c2 <- card("10", "♦")
hand <- new_blackjack_hand(c(c1, c2))

card_rank(hand$cards)    # Returns c("A", "10")
card_value(hand$cards)   # Returns c(11, 10)
```

*These functions ensure your game logic is **clean**, **type-safe**, and **readable**.*

**3. Extending with S3 + vctrs**

All relevant behaviors—printing, type coercion, equality—are registered explicitly via `vctrs::s3_register()`.

```{r eval=FALSE}
# Custom print method for card objects
print.card <- function(x, ...) {
  cat(sprintf("<card: %s of %s>\n", card_rank(x), card_suit(x)))
}

# Register vctrs casting and type methods for cards
vctrs::s3_register("vctrs::vec_cast", "card.character")
vctrs::s3_register("vctrs::vec_cast", "character.card")
vctrs::s3_register("vctrs::vec_ptype2", "card.character")
vctrs::s3_register("vctrs::vec_ptype2", "character.card")
vctrs::s3_register("vctrs::vec_ptype2", "card.card")
```

*This design ensures smooth interoperation with base R vectors and custom behaviors.*

### Benefits for Advanced Users & Developers

-   **Safer code:** Bugs from invalid card or hand objects are caught early

-   **Clearer APIs:** Functions clearly expect and return game objects

-   **Easy extension:** Add new rules or card types with minimal refactoring

-   **Better debugging:** Informative errors when misuse is detected

## 🧵 Metaprogramming & Generic Interfaces

While **`wombat21`** does not heavily rely on *metaprogramming*, it demonstrates simple use of **expression evaluation** and **S3 generics** to enable flexible logic and modular game components.

### Metaprogramming in Action

A notable use of metaprogramming appears in our AI hit/stand decision logic:

```{r eval=FALSE}
while (rlang::eval_tidy(rlang::expr(!!current_score < 17))) {
  ...
}
```

This pattern lets us **evaluate conditional logic dynamically**, which could be extended to support more advanced decision trees or strategy DSLs.

Although simple here, this approach allows:

-   Dynamic game strategies (e.g., thresholds that are formula-based)

-   Pluggable control flows without hardcoding

-   Compatibility with user-defined logic in future extensions

### Interface Design via S3

Beyond that, **interface extensibility** in `wombat21` is achieved through clean **S3 generics**:

-   `card_rank()`, `card_value()`, and `calculate_score()` all support **class dispatch**

-   Functions like `card_rank.card()` and `card_rank.character()` make logic reusable across both S3 `card` objects and plain characters

This gives:

-   **Clear separation** between internal logic and user-visible objects

-   **Easy extension** of new types (e.g., `joker_card`, `double_card`)

-   Consistent API behavior with helpful fallbacks and method dispatch

### Summary

::: {style="width:100%"}
| Aspect              | Technique Used           | Purpose                                     |
|---------------------|--------------------------|---------------------------------------------|
| Hit/stand condition | `rlang::eval_tidy()`     | Dynamically evaluate game thresholds        |
| Rank/value logic    | S3 generic dispatch      | Allow multiple input types (`card`, `char`) |
| Extendability       | `vctrs::new_rcrd()` + S3 | Build and validate extensible game types    |
:::

### Want to do more?

-   For more advanced *metaprogramming*, consider:

-   User-defined strategy expressions via quosures

-   Parameterized hit/stand rules embedded in `play()` flow

-   Callable logic passed in as closures (e.g., `split_func`, `insurance_func` – which you already support)

## ⚡ Boosting Performance with Rcpp

To speed up critical logic like **score calculation**, `wombat21` uses a C++ backend via `Rcpp`. This ensures fast, reliable, and vectorized performance even during **high-volume simulation**.

### What’s Accelerated

The most computationally intense operation in Blackjack is score calculation—especially due to the **variable value of Aces**. Instead of handling this in pure R every round, we offload it to C++:

```{r eval=FALSE}
calculate_score(hand)
```

internally calls:

```{r eval=FALSE}
calculate_score_cpp(ranks)
```

### How It Works

```{r eval=FALSE}
# R wrapper (R/score_cal.R)
calculate_score <- function(hand) {
  ...
  ranks <- card_rank(cards)
  calculate_score_cpp(ranks)
}
```

```{r cpp-example, eval=FALSE}
# C++: src/calculate_score.cpp
# int calculate_score_cpp(std::vector<std::string> ranks);
```

This ensures:

-   **Low latency** even with thousands of hands

-   **Correct Ace handling** (e.g., dual 1/11)

-   **Seamless integration** with S3/vctrs objects

### When to Use

You don’t need to call the C++ version manually—`calculate_score()` handles this for you. But advanced users can test or use it directly:

```{r eval=FALSE}
calculate_score_cpp(c("A", "A", "9"))  # Returns 21
```

**Speed Comparison (10,000 hands)**

::: {style="width:100%"}
| Method                       | Elapsed Time (sec) | Relative Speed |
|------------------------------|--------------------|----------------|
| Pure R (`calculate_score_r`) | 0.28               | 1× (baseline)  |
| Rcpp C++ (`calculate_score`) | 0.06               | 4.7× faster    |
:::

## ✅ Testing and Data Persistence

The package uses unit tests (via testthat) and persistent storage (e.g., `save_players()`/`load_players()`) to ensure reliability and reproducibility for all features.

## 🚧 Limitations

-   **wombat21** is designed for flexibility and extensibility, but there are still some current limitations:

-   **Single-deck only:** Only single-deck Blackjack is implemented; multi-deck or shoe games are not directly supported.

-   **Standard rules only:** Nonstandard Blackjack variants (e.g., double exposure, Spanish 21) are not included but could be added via user extensions.

-   **Strategy granularity:** Simulation strategies are customizable via function parameters, but built-in logic is intentionally simple.

-   **No GUI:** The package is console-based; no graphical interface is provided (though plots are interactive).

-   **Persistence:** Player data is saved locally, not in a centralized/cloud-based system.

-   **Performance:** Scoring uses C++ for speed, but some highly complex batch simulations may be slower than pure C++/Python.

## 🔄 Alternative Designs

We considered several alternative approaches in developing **wombat21**:

-   **Pure S3 vs vctrs:** Adopted vctrs for safer extensible objects, could have used pure S3.

-   **R-only vs C++ backend:** Chose C++ for batch speed; R only is simpler but slower.

-   **Functional vs OOP:** Prioritized functional interfaces over OOP for composability.

-   **Integrated GUI:** Focused on core logic, left out Shiny/desktop GUI for clarity.

-   **Database/data.table support:** Used R lists/local files for simplicity; future work could add database integration.

## 🤖 AI Statement

In compliance with University policy on responsible AI use, we openly and clearly acknowledge **extensive use of generative AI tools—including OpenAI ChatGPT and GitHub Copilot—in the development of this assignment.**

### How AI Was Used

-   **Initial code generation and scaffolding:**\
    Many basic functions, S3 class constructors, and control-flow templates (such as for card creation, game setup, simulation loops, and some unit tests) were generated with ChatGPT, then thoroughly reviewed, modified, and integrated by the team.

-   **Language and documentation refinement:**\
    AI was used to polish English language in the README, function documentation, and code comments.

-   **Debugging and best practices:**\
    We consulted ChatGPT for error explanations and debugging advice, always reviewing and adapting outputs before inclusion.

-   **Design discussion and alternatives:**\
    AI was used as a “second opinion” in discussions about object systems, file structure, and testing coverage, with all final design decisions made by the team.

### How AI Was **Not** Used

-   AI was **never** used to submit unreviewed or unedited answers.

-   No material was copied blindly from AI outputs—every section was checked, rewritten, and adapted.

-   All decisions and code integrations reflect our own understanding and academic responsibility.

### Accountability and Academic Integrity

-   We accept full responsibility for all code, documentation, and design decisions.

-   AI was used as an assistant for efficiency and idea generation, not as a replacement for our learning.

-   All AI use is fully documented here, following University transparency and integrity policies.

-   We are prepared to discuss or demonstrate any part of our work as evidence of our engagement and understanding.

### Limitations and Reflection

-   AI-generated code often required significant rewriting or adaptation to meet our standards.

-   Some AI suggestions were inefficient or incomplete; we improved or replaced these sections as needed.

-   Test coverage currently emphasizes core logic; auxiliary features may be less robust.

### Example Prompts Used

-   "Write an S3 constructor for a card object in R."

-   "How to generate a unit test for splitting hands in Blackjack using testthat?"

-   "Polish the English in this README section."

-   "Explain how to fix this vctrs error in an R test."

-   "Suggest a yaml grouping for pkgdown user-facing and internal functions."

### Statement of Responsibility

We affirm that:

-   All team members participated in using and reviewing AI-assisted code and documentation, and all share responsibility for the submitted work.

-   AI was used only as an aid; all logic, structure, and integration reflect our own skills and decision-making.

-   Our use of AI aligns with University expectations for responsible, transparent, and ethical assessment practice.

::: {style="text-align: right;"}
Prepared by: Weizhu Chen, Sun Yan Joanna Ma, Aneesh Agarwal\
Date: 2025-06-06
:::

## 🧑‍🤝‍🧑 Team Contribution Summary

::: {style="width:100%"}
+----------------+-----------------------------------------------------------------------------------------+
| Name           | Key Contributions                                                                       |
+================+=========================================================================================+
| Weizhu Chen    | \- Designed and implemented main game loop and core mechanics\                          |
|                | - Developed S3 generics/methods (e.g., `card_rank`, `vec_cast`)\                        |
|                | - Modularized and refactored codebase\                                                  |
|                | - Added simulation feature and usage examples\                                          |
|                | - Improved documentation structure and naming consistency                               |
+----------------+-----------------------------------------------------------------------------------------+
| Joanna Ma      | \- Developed multi-player and computer player support\                                  |
|                | - Implemented betting and coin management\                                              |
|                | - Created visualization functions for player ranking and coin history\                  |
|                | - Debugged and enhanced S3/vctrs compatibility\                                         |
|                | - Helped with data storage and user interface troubleshooting                           |
+----------------+-----------------------------------------------------------------------------------------+
| Aneesh Agarwal | \- Developed and integrated C++ backend for fast hand scoring (`calculate_score_cpp`)\  |
|                | - Provided function examples and unit tests\                                            |
|                | - Contributed to vignette and design documentation (including metaprogramming choices)\ |
|                | - Assisted in troubleshooting and code review                                           |
+----------------+-----------------------------------------------------------------------------------------+
:::

::: {style="text-align: right;"}
For full details of team member roles, see the [project repository](https://github.com/MonashARP/game-package-wombats).
:::

## 🤝 Conclusion & Contribution

We welcome feature requests, bug reports, and community contributions. Feel free to open an issue or submit a pull request to help make `wombat21` even better!
